---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
title: "Bivariate Exercises: 1-4"
output: html_notebook
---
## Course: Visual Analytics for Policy and Management

## Bivariate Exercises 1-4
## By Kimberly Pearson
## Version 2 

<a id='ex1'></a>
<span style="color:red"> Exercise 1:<br> Turn the bars into lollipop plot with the right components.
</span>

```{r data, eval=TRUE}
link="https://github.com/EvansDataScience/data/raw/master/crime.RData"
load(file = url(link))

titleText="What crimes happen at what times of day in Seattle?"
sourceText="Source: Seattle Open Data Portal"
```

```{r}
names(crime)
```


```{r table1, eval=TRUE}
(CrimeTotal=table(crime$crimecat,crime$Occurred.DayTime))

library(magrittr)
(CrimeTotal=table(crime$crimecat,crime$Occurred.DayTime)%>% #create table and then to the result...
        prop.table() %>% #compute proportion and then once you have the result...
        "*"(100)%>% # multiply by 100 and then...
        round(2) #...round to to decimals
        )

str(CrimeTotal)

df.T=as.data.frame(CrimeTotal) # table of proportion based on total
colnames(df.T)=c('Crime','Daytime','Percent') 

```

```{r car, eval=TRUE}
library(car)
crime$crimecat<-recode(crime$crimecat, "'BURGLARY'='BURGLARY'; 'CAR PROWL'='CAR PROWL'; 'THEFT'='THEFT'; else='OTHER'")

```

```{r table2, eval=TRUE}
table(crime$crimecat)
```


```{r lolliplot, eval=TRUE}
library(ggplot2)
base  = ggplot(df.T, aes(x = reorder(Crime, Percent), y = Percent ))
lolliplot1 = base + geom_segment(aes(y = 0, 
                                   x = Crime, 
                                   yend = Percent, 
                                   xend = Crime), color = "grey50") 
lolliplot2 = lolliplot1 + geom_point()
lolliplot3 = lolliplot2 + facet_wrap( ~ Daytime ,nrow = 1)  
lolliplot4 = lolliplot3 + coord_flip() + labs(title=titleText, subtitle=sourceText, caption="Other includes Aggravated Assault, Arson, Disorderly Conduct, DUI, 
                                              Gambling, Homicide, Narcotic, Pornography, Rape, Robbery, and 7 Other Crimes", x="Crimes", y="Percent")
lolliplot5 = lolliplot4 + theme(plot.title = element_text(hjust=0.5), plot.subtitle = element_text(hjust=0.5), axis.text.y = element_text(size=5,angle = 0)) 
lolliplot5 


```
Exercise 2: Change the heatplot to ascending order, where intensity goes from yellow to purple.

```{r heatplot, eval=TRUE}
base  = ggplot(df.T, aes(x = Daytime, y = reorder(Crime, - Percent), fill = Percent)) 
heat1 = base +  geom_tile() #heatplot
heat2 = heat1 + scale_fill_gradient(low = "yellow", 
                                   high = "purple")
heat3 = heat2 + theme_classic() 
heat4 = heat3 + theme(axis.text.x = element_text(angle = 90, vjust = 0.6), 
                      legend.title = element_blank(), #no title for legend
                      legend.position="top", 
                      legend.direction="horizontal",
                      legend.key.width=unit(1, "cm"),
                      legend.key.height=unit(1, "cm")) 
heat4 + labs(title = "Theft is the Most Popular Crime, Occurring Most in the Afternoon", subtitle = sourceText, caption="Other includes Aggravated Assault, Arson, Disorderly Conduct, DUI, 
                                              Gambling, Homicide, Narcotic, Pornography, Rape, Robbery, and 7 Other Crimes", x="Crime", y="Time of Day")
```

#Exercise 3: Complete the information needed in the previous plots.

```{r CrimeReportData, eval=TRUE}
# Crimes reported after a year 
crimeY2000=crime[(crime$DaysToReport>=365) & (crime$year>=2000),] # Crimes after 2000 that took more than a year to report 

crimeY2000$YearsToReport=crimeY2000$DaysToReport/365 # Converting the factor to be expressed in years, not days 

crimeYearGG=crimeY2000[complete.cases(crimeY2000$YearsToReport),] # Keeping only complete cases 

# Crimes reported after less than a year 
crimeWeek=crime[(crime$DaysToReport<365) & (crime$year>=2000),] # Crimes after 200 that took less than a year to report 

crimeWeek$WeeksToReport=crimeWeek$DaysToReport/7 # Converting the factor to be expressed in weeks 

crimeYearGG2=crimeWeek[complete.cases(crimeWeek$WeeksToReport) &complete.cases(crimeWeek$crimecat),] # Keeping only complete cases 

```

Below is the plot for crimes reported after one year, organized by type of crime. 

```{r CrimeYear, eval=TRUE}
base= ggplot(crimeYearGG,
             aes(x = reorder(crimecat, YearsToReport, FUN = max), # reorder! function = max says reorder based on an ascending Maximum 
                 y=YearsToReport)) 
mins = base + stat_summary(fun.y=min,size=1,
                           geom="line", linetype='dashed',show.legend = T,
                           aes(group=1,col='Min'))
minsMaxs= mins + stat_summary(fun.y=max,
                              geom="line",size=1,show.legend = F,
                              aes(group=1,col='Max'))
minsMaxsMd= minsMaxs + stat_summary(fun.y=median, size=2,
                                    geom="line",
                                    aes(group=1,col='Median'))
all3=minsMaxsMd + scale_colour_manual(name="Trends",
                                      values=c("blue", "black","red"))
all3 + coord_flip() + labs(title="Other Crimes Are Reported Later More Frequently", subtitle=sourceText, caption="Other includes Aggravated Assault, Arson, Disorderly Conduct, DUI, 
                                              Gambling, Homicide, Narcotic, Pornography, Rape, Robbery, and 7 Other Crimes", x="Crimes", y="Years to Report")
```

The next plot is of crimes that took longer than a week to report, organized by crime. 

```{r CrimeWeeks, eval=TRUE}
base = ggplot(crimeYearGG2,
              aes(x = reorder(crimecat, WeeksToReport, FUN = max),
                  y=WeeksToReport)) 
mins = base + stat_summary(fun.y=min,size=1,
                           geom="line", linetype='dashed',show.legend = T,
                           aes(group=1,col='Min'))
minsMaxs= mins + stat_summary(fun.y=max,
                              geom="line",size=1,show.legend = F,
                              aes(group=1,col='Max'))
minsMaxsMd= minsMaxs + stat_summary(fun.y=median,size=2,
                                    geom="line",
                                    aes(group=2,col='Median'))
all3=minsMaxsMd + scale_colour_manual(name="Trends",
                                      values=c("blue", "black","red"))
all3+coord_flip() + labs(title="Crimes Reported Less Than A Year from Date of Crime", subtitle=sourceText, caption="Other includes Aggravated Assault, Arson, Disorderly Conduct, DUI, 
                                              Gambling, Homicide, Narcotic, Pornography, Rape, Robbery, and 7 Other Crimes", x="Crimes", y="Weeks to Report")
```

Exercise 4: Complete the elements missing in the scatterplot of proportion of crimes by neighborhood and days to report crimes by neighborhood. 

```{r byNeigh, eval=TRUE}
daysByNeigh=tapply(crime$DaysToReport, crime$Neighborhood, mean,na.rm=T)

crimesByNeigh=tapply(crime$crimecat, crime$Neighborhood, length)%>%      
                     prop.table()%>%
                     "*"(100)%>% 
                     round(2) 
crimesByNeigh=tapply(crime$crimecat, crime$Neighborhood, length)%>%      
                     prop.table()%>%
                     "*"(100)%>% 
                     round(2) 
head(crimesByNeigh)


```

```{r transpose, eval=TRUE}
library(tibble)
as.data.frame(daysByNeigh)%>%rownames_to_column()
```

```{r merge, eval=TRUE}
daysByNeigh=as.data.frame(daysByNeigh)%>%rownames_to_column()
crimesByNeigh=as.data.frame(crimesByNeigh)%>%rownames_to_column()

num_num=merge(daysByNeigh,crimesByNeigh) # 'row name' is the "key"
head(num_num)
```

```{r scatter, eval=TRUE}
library(ggrepel)
base = ggplot(num_num, aes(daysByNeigh,crimesByNeigh,label=rowname)) 
plot1= base +  geom_point() 
plot2= plot1 + geom_text_repel(aes(label=ifelse(crimesByNeigh>=5, 
                                         num_num$rowname, ""))) 
plot2
```

```{r hexbin, eval=TRUE}
base = ggplot(num_num, aes(daysByNeigh,crimesByNeigh)) 
scatp1 = base +  geom_hex(bins = 10)
scatp2= scatp1 + geom_text_repel(aes(label=ifelse(crimesByNeigh>=5,
                                                  num_num$rowname,
                                                  ""))) 
scatp2 + scale_fill_distiller(palette ="OrRd",direction=1) + labs(title = "Downtown Seattle Experiences More Crimes But Reports Quickly", subtitle = sourceText, x = "Days to Report", y = "Perecentage of Crimes")
```



```{r Scatterplot, eval=TRUE}
base = ggplot(num_num, aes(daysByNeigh,crimesByNeigh)) 
scatp1 = base +  stat_density_2d(aes(fill = ..density..), 
                                 geom = "raster", contour = FALSE)
scatp2=scatp1+geom_text_repel(aes(label=ifelse(crimesByNeigh>=5,
                                               num_num$rowname, "")))
scatp3 = scatp2 +  theme(legend.position='none') 
scatp4= scatp3 + scale_fill_distiller(palette="Blues", direction=1) 
scatp5 = scatp4 +  scale_x_continuous(expand = c(0, 0)) + 
         scale_y_continuous(expand = c(0, 0)) 
scatp5 + labs(title = "Downtown Seattle Experiences More Crimes But Reports Quickly", subtitle = sourceText, x = "Days to Report", y = "Perecentage of Crimes")
```

